---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bacontrees

<!-- badges: start -->
[![R-CMD-check](https://github.com/Freguglia/bacontrees/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Freguglia/bacontrees/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/Freguglia/bacontrees/graph/badge.svg)](https://app.codecov.io/gh/Freguglia/bacontrees)
<!-- badges: end -->

`bacontrees` is an R package for modelling discrete sequential data using
**Ba**yesian **Con**text **Trees** (BaCon Trees). Context trees, also known as
Variable Length Markov Chains (VLMCs), are parsimonious Markov models where
the order of dependence can vary with the observed past. `bacontrees` provides:

- A **generic tree structure** (`ContextTree`) that exposes the full context
  tree for building custom algorithms.
- **Exact Bayesian inference** (`baConTree`): closed-form marginal likelihood,
  exact MAP tree, exact prior/posterior probabilities for any tree, and exact
  sampling from the posterior, all via an efficient recursive algorithm.
- A **frequentist** estimator: `fit_vlmc()` estimates the context tree via the
  context algorithm with likelihood-ratio pruning.
- Simulation utilities (`rvlmc()`).

## Installation

You can install the development version of bacontrees from
[GitHub](https://github.com/Freguglia/bacontrees) with:

```r
# install.packages("pak")
pak::pak("Freguglia/bacontrees")
```

Or with **devtools**:

```r
# install.packages("devtools")
devtools::install_github("Freguglia/bacontrees")
```

## Background

A **Variable Length Markov Chain** (VLMC) is a Markov model where the relevant
past (the *context*) that determines the next symbol's distribution has a
variable length. Further pasts are only distinguished when the data
supports it, making VLMCs much more parsimonious than fixed-order Markov chains.

The set of contexts forms a complete subtree of the full (maximal) suffix tree
over the alphabet. `bacontrees` represents this subtree explicitly as a
`ContextTree` object, whose *active* leaf nodes are exactly the contexts of the
model.

Under the Bayesian formulation, a Dirichlet prior is placed on the
transition probabilities at each context and a prior distribution over tree
structures is defined through per-node weights. The resulting posterior over
trees can be computed **exactly**: a single bottom-up recursion over the maximal
tree yields the marginal likelihood, the posterior probability of every
candidate tree, the MAP tree, and an exact sampler.

## Usage

### Included data

The package ships with two example datasets: `abc_vec` (a single sequence of
length 1000) and `abc_list` (a list of three sequences of length 1000 each),
generated from a three-symbol VLMC with alphabet `{a, b, c}`.

```{r load}
library(bacontrees)
data(abc_list)
head(abc_list[[1]], 20)
```

### Simulating sequences from a VLMC

Use `rvlmc()` to generate synthetic sequences given a context list and
the corresponding transition probability vectors.

```{r rvlmc}
alphabet      <- c("a", "b", "c")
context_list  <- c("*.a", "*.b", "*.c.a", "*.c.b", "*.c.c")
context_probs <- list(
  c(0.10, 0.20, 0.70),  # after 'a'
  c(0.33, 0.33, 0.34),  # after 'b'
  c(0.20, 0.10, 0.70),  # after 'ac'
  c(0.01, 0.98, 0.01),  # after 'bc'
  c(0.40, 0.40, 0.20)   # after 'cc'
)

set.seed(42)
seq1 <- rvlmc(1000, alphabet, context_list, context_probs)
head(seq1, 20)
```

### Frequentist fitting — `fit_vlmc()`

`fit_vlmc()` implements the classical context algorithm: build a maximal tree,
then prune nodes whose likelihood-ratio test statistic falls below `cutoff`.

```{r fit_vlmc}
fit <- fit_vlmc(abc_list, cutoff = 20, max_length = 4)
fit$getActiveNodes()
```

### Visualising a context tree

`ContextTree` (and `baConTree`) objects have an S3 `plot()` method that uses
**ggraph** internally and can be customised like any ggplot2 object.

```{r plot, fig.alt = "Context tree plot"}
plot(fit)
```

### Bayesian inference — `baConTree`

`baConTree` extends `ContextTree` with Bayesian machinery. On construction it
runs a bottom-up recursion that precomputes the normalising constants
at every node of the maximal tree. These values make all key
Bayesian quantities available as direct, closed-form computations.

```{r baConTree_init, eval = FALSE}
bt <- baConTree$new(abc_list, maximalDepth = 3, alpha = 0.01,
                    priorWeights = function(node) exp(-node$getDepth() / 3))
```

#### Marginal likelihood

The exact log marginal likelihood of the data (integrating over all tree
structures and all transition probability vectors) is available immediately
after construction:

```{r marginal, eval = FALSE}
bt$getMarginalLikelihood(log = TRUE)
```

#### MAP tree

The MAP tree is found exactly in a single top-down pass, no sampling required:

```{r map, eval = FALSE}
bt$activateMap()
bt$getActiveNodes()
```

#### Exact posterior probabilities

The prior and posterior probability of *any* candidate tree can be evaluated
exactly by activating that tree and calling `activeTreeProbabilities()`:

```{r probabilities, eval = FALSE}
bt$activateFromContexts(c("*.a", "*.b", "*.c.a", "*.c.b", "*.c.c"))
bt$activeTreeProbabilities()
```

#### Exact sampling

Draw independent samples directly from the prior or posterior without MCMC:

```{r sample, eval = FALSE}
set.seed(1)
bt$sampleTree("posterior")
bt$getActiveNodes()
```

### Metropolis-Hastings sampling — `metropolis_vlmc()`

For settings where exact enumeration is not the goal (e.g., diagnosing
multimodality or studying the mixing behaviour of MCMC on tree spaces),
`metropolis_vlmc()` provides a Metropolis-Hastings sampler over context trees.

```{r metropolis_vlmc}
library(progressr)

with_progress({
  result <- metropolis_vlmc(
    abc_list,
    n_steps       = 2000,
    max_depth     = 3,
    alpha         = 0.01,
    context_weights = function(node) exp(-node$getDepth() / 3),
    burnin        = 200
  )
})

print(result)
```

The returned object contains:

| Field        | Description                                                  |
|--------------|--------------------------------------------------------------|
| `$df`        | Data frame of context sets ranked by posterior probability   |
| `$codes`     | Named list mapping tree codes to their context vectors       |
| `$chain`     | Full MCMC chain of tree codes (for convergence diagnostics)  |
| `$baConTree` | The `baConTree` object used internally                       |

## Class reference

### `ContextTree`

The core data structure. Manages the maximal suffix tree, tracks which nodes are
*active* (the current context set), and handles data attachment. Designed to be
subclassed or used directly for building custom algorithms over context trees.

| Method | Description |
|--------|-------------|
| `$new(dataset, maximalDepth, alphabet)` | Construct a context tree |
| `$root()` | Return the root node |
| `$validate()` | Check that the maximal tree structure is valid |
| `$getMaximalDepth()` | Return the maximal depth of the tree |
| `$getActiveNodes()` | Return paths of active (leaf) nodes |
| `$getLeaves()` | Return paths of maximal-tree leaves |
| `$getInnerNodes()` | Return inner nodes of the active tree |
| `$getGrowableNodes()` | Return active nodes that can be grown |
| `$getPrunableNodes()` | Return active nodes that can be pruned |
| `$nodeExists(path)` | Check if a node with the given path exists |
| `$getParentNode(path)` | Return the parent of a node |
| `$getChildrenNodes(path)` | Return the children of a node |
| `$getSiblingNodes(path)` | Return the siblings of a node |
| `$activateRoot()` | Set active tree to root-only |
| `$activateMaximal()` | Set active tree to all maximal leaves |
| `$activateByCode(code)` | Restore a previously stored tree code |
| `$activateFromContexts(contexts)` | Set active tree to match a context vector or string |
| `$growActive(path)` | Grow the active tree at a given node |
| `$pruneActive(path)` | Prune the active tree at a given node |
| `$setData(dataset)` | Attach sequence data and compute counts |
| `$getDataset()` | Retrieve attached `Sequence` object |
| `$getAlphabet()` | Retrieve the `Alphabet` object |
| `$activeTreeCode()` | Compact binary encoding of the active tree |
| `$igraph()` | Convert to an `igraph` object |
| `plot(ct)` | Plot with **ggraph** |

### `baConTree`

Extends `ContextTree` with Bayesian inference.

| Method | Description |
|--------|-------------|
| `$new(data, maximalDepth, alpha, priorWeights, initialTree)` | Construct a Bayesian context tree and precompute recursive quantities |
| `$getMarginalLikelihood(log)` | Exact marginal likelihood of the data |
| `$activateMap()` | Set active tree to the exact MAP tree |
| `$activeTreeProbabilities(log)` | Exact prior and posterior probabilities of the active tree |
| `$sampleTree(type)` | Draw an exact independent sample from the prior or posterior |
| `$runMetropolisHastings(steps)` | Run a Metropolis-Hastings MCMC sampler (optional) |
| `$getChain()` | Retrieve the MCMC chain as a data frame |

